Abstract

### 2025-02-16 output/Main_text.Rmd
Correct grammar and errata from Word and Grammarly

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in @reckase_difficulty_1985
---a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
Third, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.
Finally, it is worth highlighting that
this model generally assumes *within-item multidimensionality*,
which will be present whenever more than one component of $`r DISCR_VECTOR`$
is non-null.
However, the model can also account for *between-item multidimensionality*,
if the $`r DISCR_VECTOR`$ parameter only has one non-null component.

@@

Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
express the items coordinates in terms of their multidimensional parameters.
However, it is worth noting that the latent space Gram matrix
also plays a role in computing these coordinates,
and will therefore have an effect on the representation.
Of course,
from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
it is easy to express the coordinates in terms of the parameters in the original
model formulation.
Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.
Of course, it is also directly proportional to
the intercept parameter $`r INTERCEPT_PARAM`$.
This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
regardless of the discrimination.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

The latter effect can be better noticed in another case, in which Reckase
uses the same item parameters to demonstrate the over-specification of
dimensions (see p. 183).
In that example, he uses a covariance matrix that practically has rank 1,
implying that the (alleged) three dimensions collapse into a single one, being
thus the test practically unidimensional.
In such a case, the covariance-based direction angles
are extremely close to 0 for all items, as one would expect
from a unidimensional test, where all items are strictly aligned with the one
dimension they measure. The reader interested can check out the parameters
computed in that case in the table in Online Resource 1.

@@

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is relevant in applications where
potentially different bases are involved.
Furthermore, not that, when using the covariance-based version formulation,
the resulting norm in the latent space is the Mahalanobis distance
[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
thus taking the latent covariance structure into account. 
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad-hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

>>>>

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in @reckase_difficulty_1985
---a decrease in the probability of a correct response
with an increase in ability makes little sense.
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
Third, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.
Finally, it is worth highlighting that
this model generally assumes *within-item multidimensionality*,
which will be present whenever more than one component of $`r DISCR_VECTOR`$
is non-null.
However, the model can also account for *between-item multidimensionality*,
if the $`r DISCR_VECTOR`$ parameter only has one non-null component.

@@

Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
express the item coordinates in terms of their multidimensional parameters.
However, it is worth noting that the latent space Gram matrix
also plays a role in computing these coordinates,
and will therefore have an effect on the representation.
Of course,
from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
it is easy to express the coordinates in terms of the parameters in the original
model formulation.
Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forward or backward with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.
Of course, it is also directly proportional to
the intercept parameter $`r INTERCEPT_PARAM`$.
This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
regardless of the discrimination.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version, however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

The latter effect can be better noticed in another case, in which Reckase
uses the same item parameters to demonstrate the over-specification of
dimensions (see p. 183).
In that example, he uses a covariance matrix that practically has rank 1,
implying that the (alleged) three dimensions collapse into a single one, being
thus the test unidimensional in practice.
In such a case, the covariance-based direction angles
are extremely close to 0 for all items, as one would expect
from a unidimensional test, where all items are strictly aligned with the one
dimension they measure. The reader interested can check out the parameters
computed in that case in the table in Online Resource 1.

@@

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is relevant in applications where
potentially different bases are involved.
Furthermore, not that, when using the covariance-based version formulation,
the resulting norm in the latent space is the Mahalanobis distance
[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
thus taking the latent covariance structure into account. 
Given these considerations,
we propose adopting the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad-hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).
